---
output:
  word_document: default
  html_document: default
---
## Assignment 5 --- Regression models: ARMAX and RegARIMA

Author: Anna (Yuanyuan) Qin
UCID: 43865058
Date: Oct.20, 2023

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(forecast)
library(tseries)
library(seasonal)
library(zoo)
library(ggplot2)
library(readr)
library(stats)
library(lubridate)
library(astsa)
library(seasonalview)
```

#### Load monthly NZ visitor data
Transform it into a time series and then plot it.

```{r}
data <- read_csv("Visitor_Monthly.csv", show_col_types = FALSE)
head(data)
tail(data)

# Create a time series object
data$Date <- as.Date(data$Date, format = "%Y-%m-%d")
data$Year <- year(data$Date)
data$Month <- month(data$Date)

ts_data <- ts(data$Actual_Counts, start = c(1980, 1), frequency = 12)

ts.plot(ts_data,col="purple",lwd=1)

start(ts_data)
end(ts_data)
```
Split visitor time serial data into 3 sets:

1. Serial I: 2020/04 - 2023/04  (ARMAX)

2. Serial II: 2000/01 - 2023/04 (RegARIMA: X-13ARIMA-SEATS)

3. Validation serial: 2023/05, 2023/06 and 2023/07


```{r}

ts1 <- window(ts_data, start = c(2020, 4), end = c(2023, 4))
ts.plot(ts1,col="darkred",lwd=1)
head(ts1)
tail(ts1)
length(ts1)

```
```{r}

ts2 <- window(ts_data, start = c(1980, 1), end = c(2023, 4))
ts.plot(ts2,col="darkgreen",lwd=1)
grid(col = "lightgray", lty = "dotted")

head(ts2)
tail(ts2)

```

####Test set ( visitors of the recent 3 months )

```{r}

ts_test <- window(ts_data, start = c(2023, 5), end = c(2023, 7))
log(ts_test) 
```
#### USE ARMAX for serial I

```{r}

acf(ts1, col="red",lwd=3, lag.max = 48)
pacf(ts1, col="green",lwd=3, lag.max = 48)

```
#### ADF test 
```{r}

ts1_lg <- log(ts1)

ts.plot(ts1_lg,col="darkred",lwd=1)
grid(col = "lightgray", lty = "dotted")

acf(ts1_lg, col="red",lwd=3, lag.max = 48)
pacf(ts1_lg, col="green",lwd=3, lag.max = 48)

```
Explore with ARIMA first

```{r}

ts1_diff <- diff(ts1)

acf(ts1_diff, lag.max = 48)
pacf(ts1_diff, lag.max = 48)

```
```{r}
adf.test(ts1_lg)

```


```{r}

ts1_ARMA <- arima(ts1_lg , order=c(1,0,1))

summary(ts1_ARMA)

ts.plot(ts1_ARMA$residuals)

acf(ts1_ARMA$residuals, lag.max = 48)

pacf(ts1_ARMA$residuals,, lag.max = 48)

```

```{r}

forecast <- forecast(ts1_ARMA, h=3)

print("Log of ARIMAX forecast values are")
forecast$mean

print("**************************************")
print("Log of Test values are")
log(ts_test)

```

#### Load the 2nd time serial: NZ monthly flight amount

```{r}

data2 <- read_csv("NZ_flights.csv", show_col_types = FALSE)
head(data2)
tail(data2)

# Create a time series object
data2$Date <- as.Date(data2$Date, format = "%Y-%m-%d")
data2$Year <- year(data2$Date)
data2$Month <- month(data2$Date)

ts_data2 <- ts(data2$TOTAL_COUNT, start = c(1997, 9), frequency = 12)

ts.plot(ts_data2,col="purple",lwd=1)

start(ts_data2)
end(ts_data2)
 
```
#### create the external factor time serial of monthly flight counts

```{r}

ts_flight <- window(ts_data2, start = c(2020, 4), end = c(2023, 4))
head(ts_flight)
tail(ts_flight)
e_flight_factor = log(ts_flight)

ts_flight_n <- window(ts_data2, start = c(2023, 5), end = c(2023, 7))
ts_flight_n

e_pre_factor <- log(ts_flight_n)

```
#### understand the correlation between time serial between flight and visitors

```{r, warning = FALSE}

lag2.plot(ts_flight, ts1, 12)

```
#### ts1 has the stronggest correlation coefficient with Lag0 of ts_flight

#### create one plot of log(ts1) and log(ts_flight). Log transformation will put the data in the similar scale
```{r, warning = FALSE}

ts_flight_df <- data.frame(date = as.Date(time(ts_flight)), value = as.numeric(log(ts_flight)), series = "Log Flight Series")

ts1_df <- data.frame(date = as.Date(time(ts1)), value = as.numeric(log(ts1)), series = "Log Visitor Series")

ggplot() +
  geom_line(data = ts_flight_df, aes(x = date, y = value, color = series), linewidth = 1) +
  geom_line(data = ts1_df, aes(x = date, y = value, color = series), linewidth = 1) +
  labs(color = "Legend")

```

#### ARIMAX model

```{r}

ts1_ARMAX <- arima(ts1_lg , order=c(2,0,0), xreg = e_flight_factor)

summary(ts1_ARMAX)

```
#### check the residual
```{r}

tsdiag(ts1_ARMAX)

ts.plot(ts1_ARMAX$resid)

acf(ts1_ARMAX$resid, col="red",lwd=3, lag.max = 48)

pacf(ts1_ARMAX$resid, col="green",lwd=3, lag.max = 48)

```
$$
Y_t = 0.415 \times Y_{t-1} + 0.270 \times Y_{t-1} - 1.780 + 1.679 \times X_t + \varepsilon_t
$$
#### forest the next 3 values
```{r}

forecast <- predict(ts1_ARMAX, newxreg = e_pre_factor)
forecast

print("Log of ARIMAX forecast values are")
forecast$pred
print("**************************************")
print("Log of Test values are")
log(ts_test)

```

### RegARIMA - X-13ARIMA-SEATS

```{r, warning = FALSE}

checkX13()

```
#### In seas, default decomposition method is seats. add x11= "" to specify the x11 decomposition method
```{r}

ts2_lg = log(ts2)

ts.plot(ts2_lg,col="darkgreen",lwd=1)
grid(col = "lightgray", lty = "dotted")

acf(ts2_lg, col="red",lwd=3, lag.max = 72)
pacf(ts2_lg, col="green",lwd=3, lag.max = 72)

```

#### Peform the first difference to remove the trends
```{r}
ts2_lg_diff = diff(ts2_lg)

acf(ts2_lg_diff, col="red",lwd=3, lag.max = 72)
pacf(ts2_lg_diff, col="green",lwd=3, lag.max = 72)

```

```{r}
ts2_ARIMA <- arima(ts2_lg , order=c(2,1,1),seasonal=list(order=c(1,1,1),period=12))

summary(ts2_ARIMA)

tsdiag(ts2_ARIMA)

ts.plot(ts2_ARIMA$resid)

acf(ts2_ARIMA$resid, col="red",lwd=3, lag.max = 48)

pacf(ts2_ARIMA$resid, col="green",lwd=3, lag.max = 48)
```

```{r}

seas_m2 <- seas(x = ts2_lg, 
                regression.aictest = NULL, 
                regression.variables = c("seasonal"))

seas_m2
```

#### summary function is to see the ARIMA model and regression coefficients
```{r}

summary(seas_m2)

```

#### use series pacakge to output the x13-ARIMA-SEATS output tables

"fct" will provide the next 12 forecast values

```{r}
series(seas_m2, "fct")
```
####  forecast  lowerci  upperci
May 2023 11.96995 11.84861 12.09129
Jun 2023 11.93051 11.80487 12.05615
Jul 2023 12.14242 12.01992 12.26493

```{r}

log(ts_test)

```

####plot function

plot the original and adjusted serial 

```{r}

plot(seas_m2)

```

#### SI plot, detrend and shows the seasonal and irregular componets
```{r}

monthplot(seas_m2)

```

```{r}
plot(resid(seas_m2))

```


####understand teh residual (white noise)
```{r}

acf(resid(seas_m2))

pacf(resid(seas_m2))

```

#### residual(white noise) would display wide-range frequencies but the energy levels are all very low
```{r}

spectrum(resid(seas_m2))

```

#### Density is residual/white noise (normal distribution ~ (0, N))
```{r}

plot(density(resid(seas_m2)))

```

#### Q-Q plot of residual
```{r}

qqnorm(resid(seas_m2))

```

